Finite-state morphological transducers for three Kypchak languages

### 2014-03-18 kipchak-lrec-2013/kipchak-paper.tex
some minor edits, §4.4.1
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@50927 72bbbca6-d526-0410-a7d9-f06f51895060

The three languages for which transducers were developed belong to the Northwestern branch of the Turkic family, which is often referred to as the Kypchak branch.  This branch can be divided into three subbranches.\footnote{It is the belief of the authors of this paper that Kyrgyz constitutes a fourth branch.}  Kumyk is a member of the Western Kypchak group, Tatar is a member of the Northern Kypchak group, and Kazakh is a member of the Southern Kypchak group \citep[82-83]{histofturkic}.  As such, each of these three languages represents one of the three branches of Kypchak.  The geographic distribution of the languages is shown in map \ref{map1}.

@@

%FIXME: description of each one, with examples
\begin{description}
 \item{\textbf{Ambiguous and `irregular' harmony}:} In Tatar, the `yoticised' vowel letters ‹е›, ‹я›, and ‹ю› each ambiguously represent a set of sounds.  For example, ‹е› can represent the sounds of /j/ followed the sound of ‹э› (as in ‹егетләр› `boys') or ‹ы› (as in ‹еллар› `years'), and is also the non-initial orthographic variant of ‹э› (as in ‹дәресләр› `lessons').  In Kumyk, ‹ё› and ‹ю› are used intervocalically to represent front rounded vowels (as in ‹гёзлер› `eyes' and ‹гюнлер› `days'), but word-initially can represent ‹й› followed by either a front (as in ‹юреклер› `hearts', ‹ёнкюлер› `darlings') or a back (as in ‹юлдузлар› `stars', ёллар `roads') vowel.  In Kazakh, the vowels /{\qipa ɘ}/ ‹і› and /ə/ ‹ы› followed by a glide /w/ ‹у› or /j/ ‹й› are written ‹у› and ‹и›, depending on the glide; for example.  The character ‹ю› is used in turn to represent ‹й› followed by either multi-phoneme value of ‹у›.  %FIXME: good Kazakh examples 
 \item \textbf{Loan words}: In Kumyk, the character ё in Russian words is pronounced as in Russian where in native words 
   it represents the sound /ø/. These two pronunciations will select for different suffixes, the Russian pronunciation taking
   back vowel suffixes самолётлар `aeroplanes', and /ø/ taking front vowel suffixes сёзлер `words'.  %FIXME: -ия in Tatar
 \item \textbf{Acronyms and numerals}: Acronyms and numerals are challenging as they will often be pronounced out loud
   in their non-abbreviated forms, for example in Kazakh, 30-дан `from thirty', 5-тен `from 5'. The ablative
   suffix -DAn alters for the phonology, but in the numeral string there is no indication of how they should alter.
\end{description}

>>>>

The three languages for which transducers were developed belong to the Northwestern branch of the Turkic family, which is often referred to as the Kypchak branch.  This branch can be divided into three subbranches.  Kumyk is a member of the Western Kypchak group, Tatar is a member of the Northern Kypchak group, and Kazakh is a member of the Southern Kypchak group \citep[82-83]{histofturkic}.\footnote{It is the belief of the authors of this paper that Kyrgyz constitutes a fourth branch.}  As such, each of these three languages represents one of the three branches of Kypchak.  The geographic distribution of the languages is shown in map \ref{map1}.

@@

\subsubsection{Ambiguous characters}
In Tatar, the `yoticised' vowel letters ‹е›, ‹я›, and ‹ю› each ambiguously represent a set of sounds.  For example, ‹е› is the non-initial orthographic variant of ‹э› (as in ‹дәресләр› `lessons'), but can also represent /j/ followed by the phoneme of ‹э› (as in ‹егетләр› `boys') or ‹ы› (as in ‹еллар› `years').
In Kumyk, ‹ё› and ‹ю› are used intervocalically to represent front rounded vowels (as in ‹гёзлер› `eyes' and ‹гюнлер› `days'), but word-initially can represent ‹й› followed by either a front vowel (as in ‹юреклер› `hearts', ‹ёнкюлер› `darlings') or a back vowel (as in ‹юлдузлар› `stars', ‹ёллар› `roads').
In Kazakh, the vowels /{\qipa ɘ}/ ‹і› and /ə/ ‹ы› followed by a glide /w/ ‹у› or /j/ ‹й› are written ‹у› and ‹и›, depending on the glide; for example.  The character ‹ю› is used in turn to represent ‹й› followed by either multi-phoneme value of ‹у›.  For example, ‹қиюда› /qəjəwdɑ/ `in the process of chopping down' and ‹киюде› /kɘjɘwdi̯ɘ/ `in the process of getting dressed' have back- and front-harmonising values, respectively, for ‹и› and ‹ю›.

These problems can often be solved by building the context into the morphophonological rules, though the addition of new contexts and exceptions to existing contexts results in rather complex \texttt{twol} rules.

\subsubsection{Loan words}
%FIXME!!! 
In Kumyk, the character ё in Russian words is pronounced as in Russian where in native words 
   it represents the sound /ø/. These two pronunciations will select for different suffixes, the Russian pronunciation taking
   back vowel suffixes самолётлар `aeroplanes', and /ø/ taking front vowel suffixes сёзлер `words'.  %FIXME: -ия in Tatar

\subsubsection{Acronyms and numerals}
%FIXME!!!
Acronyms and numerals are challenging as they will often be pronounced out loud
   in their non-abbreviated forms, for example in Kazakh, 30-дан `from thirty', 5-тен `from 5'. The ablative
   suffix -DAn alters for the phonology, but in the numeral string there is no indication of how they should alter.
